Life in the canopy: Phylogenetic filtering and morphological canalization in epiphytic ferns

### 2019-07-18 ms/manuscript.Rmd
Hide results when asserting
Update manuscript

@@

# There were significant interaction effects between growth habit and elevation on all microclimatic variables except for minimum temperature (Table S1). Slope and intercept for variables related to temperature were very similar between epiphytic and terrestrial dataloggers, but differed more strongly for those related to humidity (Table S1, Figure \@ref(fig:climate). Minimum RH increased `r min_RH_epi_slope %>% signif(2) %>% multiply_by(100) %>% percent()` per \SI{100}{\meter} elevation for epiphytic dataloggers, but only `r min_RH_ter_slope %>% signif(2) %>% multiply_by(100) %>% percent()` for terrestrial dataloggers (linear model, $\rval^2 = \Sexpr{get_result(slopes.results, "r.squared", "min_RH")}$, $\pval = \Sexpr{get_result(slopes.results, "pval", "min_RH", TRUE)}$). The Mt. Rotui \SI{830}{\meter} plot was a clear outlier, with greater mean temperature and lower minimum RH relative to other plots at similar elevation (outliers in  \fig \ref{fig:climate} at \circa \SI{800}{\meter}). Mt. Rotui is a solitary peak located between Cook's Bay and Oponohu Bay. It is more isolated with greater exposure than the other two mountains where the rest of the plots were established, which are part of a mountain complex on the interior of the island (J. Nitta, pers. obs.).

@@

```{r pca-numbers}
# Extract summary stats for mention in text
loadd(pca_results)

@@

```{r physig-numbers}
loadd(phylosig_binary_traits)
habit_d <- phylosig_binary_traits %>% filter(trait == "habit") %>% pull(D) %>% round_t(2)
hairs_d <- phylosig_binary_traits %>% filter(trait == "hairs") %>% pull(D) %>% round_t(2)
gemmae_d <- phylosig_binary_traits %>% filter(trait == "gemmae") %>% pull(D) %>% round_t(2)
glands_d <- phylosig_binary_traits %>% filter(trait == "glands") %>% pull(D) %>% round_t(2)
morphotype_d <- phylosig_binary_traits %>% filter(trait == "morphotype") %>% pull(D) %>% round_t(2)
```

@@

```{r corr-numbers}
loadd(pic_results)
stipe_pval <- pic_results %>% filter(trait == "stipe") %>% pull(p.value) %>% signif(2)
width_pval <- pic_results %>% filter(trait == "width") %>% pull(p.value) %>% signif(2)
dissection_pval <- pic_results %>% filter(trait == "dissection") %>% pull(p.value) %>% signif(2)

loadd(correlated_evo_test)
morphotype_pval <- correlated_evo_test %>% filter(trait == "morphotype") %>% pull(pval) %>% signif(2)

@@

@@

```{r sla-numbers}
loadd(fern_traits)

# First smallest SLA: Humata_anderssonii
smallest_sla_1 <- fern_traits %>% arrange(sla) %>% slice(1) %>% pull(sla) %>% signif(3)
assert_that(fern_traits %>% arrange(sla) %>% slice(1) %>% pull(species) == "Humata_anderssonii")

# Second smallest SLA: Ctenopterella_blechnoides
smallest_sla_2 <- fern_traits %>% arrange(sla) %>% slice(2) %>% pull(sla) %>% signif(3)
assert_that(fern_traits %>% arrange(sla) %>% slice(2) %>% pull(species) == "Ctenopterella_blechnoides")

@@

```{r phylosig-cont}
loadd(phylosig_cont_traits)
phylosig_cont_traits %>%
  mutate_at(vars(kval, lambda), ~round_t(., 2)) %>%
  mutate(lambda_pval = format_pval(lambda_pval)) %>%
  column_to_rownames("trait") %>%
  kable(caption = "Phylogenetic signal in continuous traits", format = "latex", booktabs = TRUE)
```

@@

```{r phylosig-binary}
loadd(phylosig_binary_traits)
phylosig_binary_traits %>%
  mutate(D = round_t(D, 2)) %>%
  column_to_rownames("trait") %>%
  kable(caption = "Phylogenetic signal in binary traits", format = "latex", booktabs = TRUE)
```

@@

```{r pics}
loadd(pic_results)

pic_results %>%
  mutate_at(vars(estimate, std.error, statistic), ~round_t(., 2)) %>%
  mutate(p.value = format_pval(p.value)) %>%
  select(trait, n_contrasts, n_pos_con, statistic, p.value) %>%
  column_to_rownames("trait") %>%
  kable(caption = "Results of phylogenetically independent contrasts analysis of quantitative (sporophyte) traits related to epiphytic growth", format = "latex", booktabs = TRUE)
```

@@

```{r corr-evo}

loadd(correlated_evo_test)

correlated_evo_test %>%
  mutate(pval = format_pval(pval)) %>%
  mutate_if(is.numeric, ~round_t(., 1)) %>%
  column_to_rownames("trait") %>%
  kable(caption = "Pagel's (1994) test of correlated evolution between binary traits", format = "latex", booktabs = TRUE)
```

@@

```{r climate, fig.height = 4, fig.width = 6, fig.cap = "Microclimatic variables."}
loadd(climate_plot)
climate_plot
```

@@

```{r pca, fig.height = 6, fig.width = 6, fig.cap = "Trait PCA."}
loadd(pca_plot)
pca_plot
```

@@

```{r tree, fig.height = 8, fig.width = 6, fig.cap = "Traits on tree."}
loadd(traits_with_tree)
traits_with_tree
```

@@

```{r cwm-div, fig.height = 8, fig.width = 6, fig.cap = "Community-weighted means."}
loadd(combined_cwm_plots)
combined_cwm_plots
```

@@

```{r comm-div, fig.height = 8, fig.width = 6, fig.cap = "Community diversity metrics."}
loadd(combined_comm_div_plots)
combined_comm_div_plots
```

>>>>

# Load analysis results
loadd(climate_plot)
loadd(combined_comm_div_plots)
loadd(combined_cwm_plots)
loadd(comm)
loadd(correlated_evo_test)
loadd(correlated_evo_test)
loadd(diversity_data)
loadd(fern_traits)
loadd(model_summaries)
loadd(pca_plot)
loadd(pca_results)
loadd(phylosig_binary_traits)
loadd(phylosig_binary_traits)
loadd(phylosig_cont_traits)
loadd(pic_results)
loadd(t_test_results)
loadd(traits_with_tree)

@@

# There were significant interaction effects between growth habit and elevation on all microclimatic variables except for minimum temperature (Table S1). Slope and intercept for variables related to temperature were very similar between epiphytic and terrestrial dataloggers, but differed more strongly for those related to humidity (Table S1, Figure \@ref(fig:climate). Minimum RH increased `r min_RH_epi_slope %>% signif(2) %>% multiply_by(100) %>% percent()` per \SI{100}{\meter} elevation for epiphytic dataloggers, but only `r min_RH_ter_slope %>% signif(2) %>% multiply_by(100) %>% percent()` for terrestrial dataloggers (linear model, $\rval^2^ = \Sexpr{get_result(slopes.results, "r.squared", "min_RH")}$, $\pval = \Sexpr{get_result(slopes.results, "pval", "min_RH", TRUE)}$). The Mt. Rotui \SI{830}{\meter} plot was a clear outlier, with greater mean temperature and lower minimum RH relative to other plots at similar elevation (outliers in  \fig \ref{fig:climate} at \circa \SI{800}{\meter}). Mt. Rotui is a solitary peak located between Cook's Bay and Oponohu Bay. It is more isolated with greater exposure than the other two mountains where the rest of the plots were established, which are part of a mountain complex on the interior of the island (J. Nitta, pers. obs.).

@@

```{r pca-numbers}
# Extract summary stats for mention in text

@@

```{r physig-numbers}
habit_d <- phylosig_binary_traits %>% filter(trait == "habit") %>% pull(D) %>% round_t(2)
hairs_d <- phylosig_binary_traits %>% filter(trait == "hairs") %>% pull(D) %>% round_t(2)
gemmae_d <- phylosig_binary_traits %>% filter(trait == "gemmae") %>% pull(D) %>% round_t(2)
glands_d <- phylosig_binary_traits %>% filter(trait == "glands") %>% pull(D) %>% round_t(2)
morphotype_d <- phylosig_binary_traits %>% filter(trait == "morphotype") %>% pull(D) %>% round_t(2)
```

@@

```{r corr-numbers}
stipe_pval <- pic_results %>% filter(trait == "stipe") %>% pull(p.value) %>% signif(2)
width_pval <- pic_results %>% filter(trait == "width") %>% pull(p.value) %>% signif(2)
dissection_pval <- pic_results %>% filter(trait == "dissection") %>% pull(p.value) %>% signif(2)
morphotype_pval <- correlated_evo_test %>% filter(trait == "morphotype") %>% pull(pval) %>% signif(2)
```

@@

### Functional and Phylogenetic Diversity

```{r comm-div-numbers}
# Format model summaries for printing
model_summ_print <-
  model_summaries %>%
  mutate(r = round_t(r.squared, 2), 
         p = format_pval(p.value))

# Format t-test results for printing
t_test_results_print <-
  t_test_results %>%
  mutate(p = format_pval(p.value))

# Define simple function for extracting value of interest from
# diversity model results
get_mod_result <- function(resp_var_select, habit_select, 
                           indep_var_select, value) {
  model_summ_print %>% 
    filter(
      resp_var == resp_var_select, 
      habit == habit_select, 
      indep_var == indep_var_select) %>% 
    pull(!!enquo(value))
}

# Make tibble of abundances
abundances <-
  comm %>%
  mutate(abundance = 
           comm %>%
           select_if(is.numeric) %>%
           rowSums()
  ) %>%
  select(species, abundance)

# Join this with trait data
traits_with_abun <- left_join(
  fern_traits, abundances
)

# Get total number of epiphytic and terrestrial species observed in all plots
n_epi <- traits_with_abun %>% filter(abundance > 0, habit == "epiphytic") %>% 
  pull(species) %>% n_distinct
n_ter <- traits_with_abun %>% filter(abundance > 0, habit == "terrestrial") %>% 
  pull(species) %>% n_distinct

# Calculate mean richness by growth habit
mean_rich <-
  diversity_data %>%
  group_by(habit) %>%
  summarize(
    mean_rich = mean(ntaxa) %>% round_t(1),
    sd_rich = sd(ntaxa) %>% round_t(1)
  ) %>%
  mutate(stat = glue("{mean_rich} ± {sd_rich}"))
```

On a single-trait basis, distribution of community-weighted mean values (CWMs) showed significant correlation with elevation for some traits, but no significant relationship was detected for any traits with minimum RH (\fig \@ref(fig:cwm-div)). 
Specific leaf area decreased with elevation for epiphytic (linear model, \rval^2^ = `r get_mod_result("sla", "epiphytic", "el", "r")`, \pval = `r get_mod_result("sla", "epiphytic", "el", "p")`) and terrestrial (linear model, \rval^2^ = `r get_mod_result("sla", "terrestrial", "el", "r")`, \pval = `r get_mod_result("sla", "terrestrial", "el", "p")`) communities. 
Degree of lamina dissection decreased with elevation in epiphytes (linear model, \rval^2^ = `r get_mod_result("dissection", "epiphytic", "el", "r")`, \pval = `r get_mod_result("dissection", "epiphytic", "el", "p")`), but not terrestrial communities. 
Stipe length showed a small, but significant, increase with elevation for epiphytic communities (linear model, \rval^2^ = `r get_mod_result("stipe", "epiphytic", "el", "r")`, \pval = `r get_mod_result("stipe", "epiphytic", "el", "p")`). 
Rhizome diameter decreased with elevation for terrestrial communities (linear model, \rval^2^ = `r get_mod_result("rhizome", "terrestrial", "el", "r")`, \pval = `r get_mod_result("rhizome", "terrestrial", "el", "p")`), but increased slightly in epiphytes (linear model, \rval^2^ = `r get_mod_result("rhizome", "epiphytic", "el", "r")`, \pval = `r get_mod_result("rhizome", "epiphytic", "el", "p")`).

We observed a total of `r traits_with_abun %>% filter(abundance > 0) %>% pull(species) %>% n_distinct` species across all plots, including `r n_epi` epiphytic and `r n_ter` terrestrial species. Mean species richness was `r mean_rich %>% filter(habit == "epiphytic") %>% pull(stat)` \si{\species}per plot for epiphytes and `r mean_rich %>% filter(habit == "terrestrial") %>% pull(stat)` \si{\species} per plot for terrestrial species (not significantly different; two-sided \tval-test, \pval = `r t_test_results_print %>% filter(resp_var == "ntaxa") %>% pull(p)`; \fig \@ref(fig:comm-div)). Species richness increased linearly with elevation and minimum RH for epiphytes, but not terrestrial communities (\fig \@ref(fig:comm-div)). Trait richness (FRic) was lower in epiphytic vs. terrestrial communities (two-sided \tval-test, \pval = `r t_test_results_print %>% filter(resp_var == "FRic") %>% pull(p)`), whereas functional diversity (FDiv) was lower in terrestrial communities (two-sided \tval-test, \pval = `r t_test_results_print %>% filter(resp_var == "FDiv") %>% pull(p)`), and functional evenness (FEve) did not differ between the two. FEve decreased with elevation, but not minimum RH, in terrestrial communities (\fig \@ref(fig:comm-div)). Epiphytic communities showed a trend of increasing trait richness along the environmental gradient, with greater FRic at high elevations (linear model, \rval^2^ = `r get_mod_result("FRic", "epiphytic", "el", "r")`, \pval = `r get_mod_result("FRic", "epiphytic", "el", "p")`) and plots with greater minimum RH (linear model, \rval^2^ = `r get_mod_result("FRic", "epiphytic", "min_RH", "r")`, \pval = `r get_mod_result("FRic", "epiphytic", "min_RH", "p")`). Epiphytic communities were more phylogenetically clustered than terrestrial communities overall when measured using MNTD (two-sided \tval-test, \pval = `r t_test_results_print %>% filter(resp_var == "mntd.obs.z") %>% pull(p)`) but not MPD (\pval = `r t_test_results_print %>% filter(resp_var == "mpd.obs.z") %>% pull(p)`; \fig \@ref(fig:comm-div)). No significant trend was detected in phylogenetic community structure with environment in either terrestrial or epiphytic communities (\fig \@ref(fig:comm-div)).

@@

```{r sla-numbers, results = "hide"}
# First smallest SLA: Humata_anderssonii
smallest_sla_1 <- fern_traits %>% arrange(sla) %>% slice(1) %>% pull(sla) %>% signif(3)
assert_that(fern_traits %>% arrange(sla) %>% slice(1) %>% pull(species) == "Humata_anderssonii")

# Second smallest SLA: Ctenopterella_blechnoides
smallest_sla_2 <- fern_traits %>% arrange(sla) %>% slice(2) %>% pull(sla) %>% signif(3)
assert_that(fern_traits %>% arrange(sla) %>% slice(2) %>% pull(species) == "Ctenopterella_blechnoides")
```

@@

```{r phylosig-cont}
phylosig_cont_traits %>%
  mutate_at(vars(kval, lambda), ~round_t(., 2)) %>%
  mutate(lambda_pval = format_pval(lambda_pval)) %>%
  column_to_rownames("trait") %>%
  kable(caption = "Phylogenetic signal in continuous traits", format = "latex", booktabs = TRUE)
```

@@

```{r phylosig-binary}
phylosig_binary_traits %>%
  mutate(D = round_t(D, 2)) %>%
  column_to_rownames("trait") %>%
  kable(caption = "Phylogenetic signal in binary traits", format = "latex", booktabs = TRUE)
```

@@

```{r pics}
pic_results %>%
  mutate_at(vars(estimate, std.error, statistic), ~round_t(., 2)) %>%
  mutate(p.value = format_pval(p.value)) %>%
  select(trait, n_contrasts, n_pos_con, statistic, p.value) %>%
  column_to_rownames("trait") %>%
  kable(caption = "Results of phylogenetically independent contrasts analysis of quantitative (sporophyte) traits related to epiphytic growth", format = "latex", booktabs = TRUE)
```

@@

```{r corr-evo}
correlated_evo_test %>%
  mutate(pval = format_pval(pval)) %>%
  mutate_if(is.numeric, ~round_t(., 1)) %>%
  column_to_rownames("trait") %>%
  kable(caption = "Pagel's (1994) test of correlated evolution between binary traits", format = "latex", booktabs = TRUE)
```

@@

```{r climate, fig.height = 4, fig.width = 6, fig.cap = "Microclimatic variables."}
climate_plot
```

@@

```{r pca, fig.height = 6, fig.width = 6, fig.cap = "Trait PCA."}
pca_plot
```

@@

```{r tree, fig.height = 8, fig.width = 6, fig.cap = "Traits on tree."}
traits_with_tree
```

@@

```{r cwm-div, fig.height = 8, fig.width = 6, fig.cap = "Community-weighted means."}
combined_cwm_plots
```

@@

```{r comm-div, fig.height = 8, fig.width = 6, fig.cap = "Community diversity metrics."}
combined_comm_div_plots
```
